State of the Practice for Lattice Boltzmann Method Software

### 2022-02-14 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions do Section 8 - Developer pain points

@@

Only so much information can be gleaned by looking at software repositories.  To
gain additional insight, we need to interview developers.  We need to learn
their concerns, how they deal with these concerns, and what pain points exist
for them.  We wish to know what practices are used by the top LBM projects, so
that others can potentially emulate these practices. We also wish to identify
new practices by borrowing successful ideas from other domains. For LBM
developers interested in future cutting edge practices, ideas that are proposed,
but not yet in practice, should also be identified.  The above points are
covered by the questions outlined below:

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on LBM software projects?
	(Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from LBM compare to the pain points
	for research software in general? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
	developers:
	\begin{enumerate} 
		\item [\rqref{RQ_Concerns}.a] what software development concerns exist; and, (Section~\ref{Sec_Concerns})
		\item [\rqref{RQ_Concerns}.b] what specific practices are taken to
		address these concerns? (Section~\ref{Sec_AddressConcerns})
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What, if any, existing research software best practice, not currently used
	by LBM software, could address the concerns identified in
	\rqref{RQ_Concerns}). (Section~\ref{Sec_AddressConcerns})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What practices
	could be employed by future LBM developers to address the pain points found
	in~\rqref{RQ_PainPoints} and the concerns identified
	in~\rqref{RQ_Concerns}.a? (Section~\ref{Sec_AddressConcerns})

@@

We requested interviews with developers for all 24 packages, except Munubi,
since Munubi was added after our ethics approved had expired.  We were able to
recruit 4 developers to participate in our study.  Our potential interview
subjects were found from project websites, code repositories, publications, and
biographic information on institution web-pages. Meeting were held using on-line
meeting software (either Zoom or Teams), which facilitated recording and
automatic transcription of the meetings.  The interview process was approved by
the McMaster University Research Ethics Board under the application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

@@

This section aims to answer the research questions: What are the pain points for
developers working on research software projects (\rqref{RQ_PainPoints})?; and,
How do the pain points of developers from LBM compare to the pain points for
research software in general (\rqref{RQ_ComparePainPoints})?

Roadmap.  Go through pain points, link to existing literature, then section on
pain points not mentioned.  Ways to address pain points in the next section

Maybe just do a table of pain points?  and then contrast the pain points with
the literature.  The mitigation strategies should be elsewhere.  Some of the
mitigation strategies have already been mentioned, like documentation.

Likely possible to combine pain points and developer concerns.  Emphasizes what
the pain points are, and how they compare to other studies.  Don't propose
solutions in this section.  Say that all of the information comes from the
interviews (as mentioned at the start of the next section).

Pain points that others, but not LBM developers, mentioned in surveys include
\citep{PintoEtAl2018}: Cross-platform compatibility, interruptions while coding,
scope bloat, lack of user feedback, mismatch between coding skills and
subject-matter skills (for instance, ``I am a mathematician, not an expert on
software''), lack of formal reward system, hard to collaborate on software
projects, aloneness.

\subsection{Lack of Development Time}

A developer of pyLBM noted that their small development team has a lack of time
to implement new features. Small development teams are common for LBM software
packages. Team members are almost always part of the same institute or already
know each other from other projects. External contributions are rare despite
many of the projects accepting them. \wss{Other projects cite the same pain
point - Wiese paper}  Aside from on-boarding new developers, time constraints
could be mitigated by increasing developer efficiency, which could be addressed
in several ways, including by improving the quality of documentation, or
incorporating automatic code generation. \wss{Potential ways to address the pain
points might go into a separate section?  Maybe propose the strategy, like code
generation, and then list the pain points that are addressed?}

Lack of time was also mentioned by \citet{PintoEtAl2018} and
\citet{PintoEtAl2016}.

\subsection{Lack of Software Development Experience}

A lack of software development experience was noted by the developer of TCLB.
Many of the team members on their project are domain experts and there can be a
steep learning curve before these members contribute good quality source code.
This problem has been mitigated by re-writing the code to more easily enable
future contributions.  On a related point, improving the software engineering
education or experience of developers is also an idea that was brought up by
several developers.  \wss{The Wiese paper and others talk about the background
of developers. The knowledge for developing research software primarily comes
from peers and through self-study, not from formal training
\citep{HannayEtAl2009}.  The thinking of the developer on the role of formal
education matches the trend \citet{PintoEtAl2018} observes, where in their
replication of a previous study \citep{HannayEtAl2009}, they show a growing
interest in formal training (From 13\% of respondants in 2009 to 22\% in 2018).}

\subsection{Lack of Incentive and Funding}

The TCLB developer noted a lack of incentive and funding in academia for
developing widely used scientific software. The importance of funding for
scientific software has been discussed in \cite{gewaltig2012quality}, which
notes that ``software tools are developed and maintained only for as long as
there is explicit or implicit funding''. The developer further commented that
there are no journals that publish such scientific software source code.
However, there are ways to get such source code cited. Work has been done to
address this in \citep{smith2016software}, which presents a set of software
citation principles and discusses ``how they could be used to implement software
citation in the scholarly community'' \citep{katz2019software}.  \wss{other
papers also mention this pain point.  Solutions proposed by Katz, journal of
open source etc.}

\subsection{Lack of External Support}

Another raised concern was that there are no organizations helping with the
development of good quality software; but some do exist, including
\href{https://bssw.io/}{Better Scientific Software (BSSw)},
\href{https://www.software.ac.uk/}{Software Sustainability Institute}
\citep{CrouchEtAl2013}, and \href{https://software-carpentry.org/}{Software
Carpentry}. Some SCS developers may not be familiar with these organizations.
\citep{WilsonAndLumsdaine2006}

Scientific software is often developed in-house by the very researchers that
temporarily use it in their own research. Empirical studies of such
``professional end-user development'' of SCS is noted in \citep{segal2007end}.
This kind of software has a defined user and purpose, and often does not meet
the standards that would be required by external users. It has been categorized
as a ``private tool'' by \citep{gewaltig2012quality}, which notes that despite
often being made freely available, ``it is not always clear that it is
sufficiently mature in terms of domain coverage, validity, documentation or
usability, to be useful to other researchers''. It is less common for such ad
hoc scientific software to become ``user-ready software'', which ``is not only
research-ready, but should have most of the attributes commonly expected of
commercial software products including broadness of scope, robustness,
demonstrable correctness and adequate documentation''
\citep{gewaltig2012quality}. As software becomes user-ready, it can become
commercialized and closed-source.  \wss{remove this paragraph?}

\wss{Remove this section?  The external support organizations will likely come
up as the mitigation for some of the other pain points.}

\subsection{Parallelization and Continuous Integration}

Setting up parallelization was also noted as a technical pain point by one of
the developers, and the introduction of continuous integration by another.
Software development knowledge, and automatic code generation, could mitigate
such pain points. As already noted, many of the developers are domain experts
and not professional software developers. The developer of TCLB noted that
eliminating equivalent statements using macros had helped improve the quality of
their source code, specifically helping with reusing code to run on both the CPU
and GPU. \wss{The point about macros isn't really related to parallelization or
continuous integration.}

\subsection{Ensuring Correctness}

Difficulties with ensuring correctness were also noted by several developers.
They indicated that tests are run on all new source code additions, testing both
individual modules and the system to verify correctness. These tests compare the
package output to known correct output using test cases. \wss{oracle}  The
developer of TCLB commented that the amount of testing data that is needed for
some cases is a problem as free testing services do not offer capabilities to
store and process such large amounts of data, and in-house testing solutions
needed to be created to address this limitation. The solution for this has been
to limit the size of the testing problems, and to run tests in small batches
with few iterations. \wss{The Weiss paper alludes this this as well.  I think
the Kelly and Sanders paper likely talks about testing problems and correctness.
``while many scientists believe that software testing is important, a smaller
number believe they have sufficient understanding about testing concepts''
\citep{HannayEtAl2009}.}

\wss{** BEGIN Testing and Verification **}

\citet{Nguyen-HoanEtAl2010} says the following:
\begin{itemize}
	\item 38 of 60 respondents performed unit tests
	\item 23 of 60 respondents performed integration tests
	\item ``In our results, the four most common activities involve comparisons
	with known, trusted, or prior data. Verification and peer review were three
	of the four least commonly performed activities.'' \wss{testing is not the
	only way to verify software.  This addresses the oracle problem.}
\end{itemize}

May do more testing than artifacts indicate, but that is still a problem.
Artifacts should be available so that others can verify the results, verify
their installation.  Part of an assurance case means that a third party can
verify the work.

\wss{** END Testing and Verification **}

``The questions of what constitutes an adequate testing oracle for scientific
software and what constitutes adequate test data are left open. Some of the
validation testing we researched was running up against the practical limits;
they were using whatever oracle they could scrape together from industry data,
benchmarks, and expert judgment.'' \citep{KellyAndSanders2008}

\wss{Seems that LBM not concerned with the oracle problem?}

A few obstacles related to users were found. Several developers noted that users
sometimes try to use incorrect LBM method combinations to solve their problems.
Furthermore, some users think that the packages will work out of the box to
solve their cases, while the packages require both a good understanding of CFD
and an understanding of the requirements for formulating problems in the
individual packages, which can be a significant endeavour. These software
packages are not like commercial software packages. They are generally set up to
solve specific research problems, and are often primarily used by their
developers. While they are modifiable to solve similar problems, these
modifications are not trivial. Better documentation, with attention on
traceability, and automatic code generation are suggested when designing
software for change, and would help with these modifications. \wss{Should decide
if want to introduce ideas for improving within the presentation of the pain
point, or separately?}  So far this problem of on-boarding new users has been
addressed by updating the documentation to better inform users of the underlying
LBM theory and package requirements. Similar issues with LBM parameters were
noted by another developer. Updating the user interface to better explain
theoretical principles, as well as test user input for compatibility, was the
implemented solution. As noted above, sometimes frequently asked questions on
the underlying theory and on how to use the software are answered in the
documentation.  The interviewed developer of ESPResSo commented that parts of
their package's source code had been refactored to Python to help address
usability issues. Python was perceived as a much more usable language, and it
would be easy for future users and developers to learn and understand the source
code.

\subsection{Technical Debt}

A few potential future obstacles were noted. The developer of ESPResSo noted
that their source code had been written with a specific application in mind and
that due to this there was too much coupling between components of the source
code. This results in technical debt, having an impact on future modifiability
and reusability when trying to extend the software, and the code would need to
be refactored.  As noted above, difficulties with ensuring future correctness
could also arise. As new methods and functionality is added into the software,
new test cases and test data will need to be developed.  \wss{Weiss mentions
mainainability related issues under pain points.}

\subsection{Quality of Documentation}

Interviewees commented that documentation is important and that its quality
could be improved. As already noted, there is often no time or funding for
maintaining quality documentation for software that is rarely used outside of
the development team. Furthermore, the documentation generally only provides a
shallow overview of the underlying CFD theory. Users would be well advised to
already be familiar with these topics, or they should spend significant time
referencing theory resources. The documentation instead generally focuses on
explaining how to use the software. It is of course not feasible for package
documentation to address the underlying physics topics in detail, so it is
advised that the package documentation links to resources that better explain
the underlying theory. Sometimes, frequently asked questions about the
underlying theory are answered in the documentation. OpenLB has an artifact for
such questions.  \wss{Weise mentions documentation.}  \wss{Partly solve the
problem by documentation user characteristics, to explicitly state what the user
is assumed to already know.}

The developer comments emphasized an importance on source code, while
documentation seems to be of secondary importance. It must be stressed that
improving documentation could benefit development and help eliminate some of the
developer concerns that were raised. The use of automatic document generation
tools that capture scientific and computing knowledge, and transform it into
software artifacts, is advised. Drasil is an automatic document generation tool
that is further discussed in Section~\ref{highlightedrecommendations}.

Poor software documentation is also mentioned by \citet{PintoEtAl2018} and
\citet{LethbridgeEtAl2003}.  The survey from \citet{LethbridgeEtAl2003} is for
software in general, not just research software.

@@

Answers \rqref{RQ_Concerns}.  This section presents qualitative findings on
software qualities as gleaned from interviewing 4 developers. The full interview
questions are found in \citet{SmithEtAl2021}.

@@

>>>>

\newcounter{pnum} %pain point number
\newcommand{\ppthepnum}{P`'\thepnum}
\newcommand{\ppref}[1]{P\ref{#1}}

@@

Only so much information can be gleaned by looking at software repositories.  To
gain additional insight, we need to interview developers.  We need to learn
their concerns, how they deal with these concerns; we need to learn what pain
points exist for them.  We wish to know what practices are used by the top LBM
projects, so that others can potentially emulate these practices. We also wish
to identify new practices by borrowing successful ideas from other domains. For
LBM developers interested in future cutting edge practices, ideas that are
proposed, but not yet in practice, should also be identified.  The above points
are covered by the questions outlined below:

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on LBM software projects?
	(Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from LBM compare to the pain points
	for research software in general? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
	developers what specific practices are taken to address the pain point
	concerns? (Section~\ref{Sec_AddressConcerns})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What, if any, existing research software best practice, not currently used
	by LBM software, could address the concerns identified in
	\rqref{RQ_Concerns}). (Section~\ref{Sec_AddressConcerns})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What practices
	could be employed by future LBM developers to address the pain points found
	in~\rqref{RQ_PainPoints} and the concerns identified
	in~\rqref{RQ_Concerns}? (Section~\ref{Sec_AddressConcerns})

@@

We requested interviews with developers for all 24 packages, except Munubi,
since Munubi was added after our ethics approved had expired.  We were able to
recruit 4 developers to participate in our study.  Our potential interview
subjects were found from project websites, code repositories, publications, and
biographic information on institution web-pages. Meeting were held using on-line
meeting software (either Zoom or Teams), which facilitated recording and
automatic transcription of the meetings.  Each meeting lasted about an hour. The
interview process was approved by the McMaster University Research Ethics Board
under the application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

@@

Based on interviews with 4 developers, this section aims to answer the research
questions: i) What are the pain points for developers working on research
software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain points of
developers from LBM compare to the pain points for research software in general
(\rqref{RQ_ComparePainPoints})?  Below we go through each of the identified pain
points and include citations that contrast the LBM experience with observations
from researchers in other domains.  Potential ways to address the pain points
are covered in Section~\ref{Sec_AddressConcerns}. The full interview questions
are found in \citet{SmithEtAl2021}.

\citep{PintoEtAl2018} lists some pain points that did not come up in our
conversations with LBM developers: Cross-platform compatibility, interruptions
while coding, scope bloat, lack of user feedback, hard to collaborate on
software projects, and aloneness. \citep{WieseEtAl2019} repeat some of the
previous pain points and add the following: dependency management, data handling
concerns (like data quality, data management and data privacy), reproducibility,
and software scope determination. Although LBM developers did not mention these
pain points, we cannot conclude that they are not relevant for LBM software
development, since we only interviewed 4 LBM developers for about an hour each.

\begin{enumerate}

	\item[P\refstepcounter{pnum}\thepnum \label{P_LackDevTime}:] \textbf{Lack of
	Development Time} A developer of pyLBM noted that their small development
	team has a lack of time to implement new features. Small development teams
	are common for LBM software packages (as shown in the measurement table
	excerpt in Figure~\ref{measurement_template_image}). Lack of time is also
	highlighted as a pain point by other research software developers
	\citet{PintoEtAl2018, PintoEtAl2016, WieseEtAl2019}.

	\item[P\refstepcounter{pnum}\thepnum \label{P_LackSoftDevExp}:] \textbf{Lack
	of Software Development Experience} A lack of software development
	experience was noted by the developer of TCLB, and others noted a need for
	improving software engineering education. Many of the team members on their
	project are domain experts, not computer scientists or software engineers.
	This same trend is noted by \citet{Nguyen-HoanEtAl2010}, which showed only
	23\% of research software survey respondents having a computing-related
	background. Similarly, \citep{UditAndKatz2017} show that the majority (54\%)
	of postdocs have not received training in software development.  The LBM
	developer suggesting an increasing role for formal software education
	matches the trend observed by \citet{PintoEtAl2018}, where their replication
	of a previous study \citep{HannayEtAl2009}, shows a growing interest in
	formal training (From 13\% of respondents in 2009 to 22\% in 2018).
	\citep{PintoEtAl2018} found that some developers feel there is a mismatch
	between coding skills and subject-matter skills.

	\item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
	Incentive and Funding} The TCLB developer noted a lack of incentives and
	funding in academia for developing widely used scientific software. This
	problem has also been noted others \citep{gewaltig2012quality, Goble2014,
	KaterbowAndFeulner2018}.  \citet{WieseEtAl2019} reported developer pains
	related to publicity, since historically publishing norms make it difficult
	to get credit for creating software.  As studied by
	\citet{HowisonAndBullard2016}, research software (specifically biology
	software, but the trend likely applies to other research software domains)
	is infrequently cited.  \citep{PintoEtAl2018} also mentions the lack of
	formal reward system for research software.

	\item[P\refstepcounter{pnum}\thepnum \label{P_LackExtSupport}:] \textbf{Lack
	of External Support} A concern was raised that there are no organizations
	helping with the development of good quality software.  This concern is not
	echoed in the literature because there are such organizations, including
	\href{https://bssw.io/} {Better Scientific Software (BSSw)},
	\href{https://www.software.ac.uk/} {Software Sustainability Institute}
	\citep{CrouchEtAl2013}, and \href{https://software-carpentry.org/}{Software
	Carpentry} \wss{WilsonAndLumsdaine2006, Wilson2016}. Over time awareness of
	these groups will grow, so this pain point is likely to disappear in the
	future for LBM and other research software developers.

	\item[P\refstepcounter{pnum}\thepnum \label{P_TechnologyHurdles}:]
	\textbf{Technology Hurdles} Technology pain points include setting up
	parallelization and continuous integration.

	\item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
	\textbf{Ensuring Correctness} Difficulties with ensuring correctness were
	noted by several developers. The TCLB developer commented that the amount of
	testing data needed was sometimes problematic, since free testing services
	do not offer adequate facilities for large amounts of data, which means
	in-house testing solutions are needed.  Other research software domains
	point to different problems with testing, as follows: i)
	\citep{PintoEtAl2018} mention the problem of insufficient testing; ii) the
	survey of \citep{HannayEtAl2009} shows that more developers think testing is
	important than the number that believe they have a sufficient understanding
	of testing concepts; and, iii) \citep{HannayEtAl2009, KanewalaAndBieman2013,
	KellyEtAl2011, WieseEtAl2019} point to the oracle problem, which
	occurs in research software when we do not have a means to judge the
	correctness of the calculated solutions.  Since LBM developers did not
	allude to the oracle problem in our conversation, we believe that they have
	developed techniques to work around the oracle problem.

	\item[P\refstepcounter{pnum}\thepnum \label{P_Usability}:]
	\textbf{Usability} Several developers noted that users sometimes try to use
	incorrect LBM method combinations to solve their problems. Furthermore, some
	users think that the packages will work out of the box to solve their cases,
	while in reality CFD knowledge needs to be applied to correctly modify the
	packages for the new endeavour.  \citep{WieseEtAl2019} also interacted with
	developers that mentioned that users do not always have the expertise
	required to install or use the software.

	\item[P\refstepcounter{pnum}\thepnum \label{P_TechDebt}:] \textbf{Technical
	Debt} The developer of ESPResSo said that their source code was written with
	a specific application in mind, which later caused too much coupling between
	components in the source code. This results in technical debt
	\citet{KruchtenEtAl2012}, which has an impact on future modifiability and
	reusability. Concern with technical debt is likely why researchers in the
	survey of \citep{Nguyen-HoanEtAl2010} rated maintainability as the third
	most important software quality. More recently the push for sustainable
	software \citet{deSouzaEtAl2019} is motivated by the pain that past
	developers have had with accumulating too much technical debt.

	\item[P\refstepcounter{pnum}\thepnum \label{P_Documentation}:]
	\textbf{Quality of Documentation} One of the consequences of pain point
	\ppref{P_LackDevTime} is sometimes inadequate documentation. The
	documentation for LBM is challenging because of the complexity of the
	underlying CFD theory and the differing backgrounds of users. Poor research
	software documentation is also mentioned by others \citep{PintoEtAl2018,
	WieseEtAl2019} and the problem also arises with non-research software
	\citep{LethbridgeEtAl2003}.  Recommendations on the development of research
	software state that developers should critically evaluate their own
	development processes in terms of quality assurance and comply with
	international standards for software documentation
	\citep{KaterbowAndFeulner2018}.

@@

@@

\wss{Address lack of development time (\ppref{P_LackDevTime}): Aside from
on-boarding new developers, time constraints could be mitigated by increasing
developer efficiency, which could be addressed in several ways, including by
improving the quality of documentation, or incorporating automatic code
generation.}

\wss{Address lack of funding (\ppref{P_LackFunding}): The developer further
commented that there are no journals that publish such scientific software
source code. However, there are ways to get such source code cited. Work has
been done to address this in \citep{SmithEtAl2016-softcite}, which presents a
set of software citation principles and discusses ``how they could be used to
implement software citation in the scholarly community''
\citep{katz2019software}. other papers also mention this pain point.
Solutions proposed by Katz, journal of open source etc. Journal of Open Source
Software \citet{SmithEtAl2018-joss}.  Calling for direct software citations
\citep{SmithEtAl2016-softcite}.  Guidance on software citation from
\citet{KatzEtAl2021}.  Discussed in CarverEtAl2021 paper.}

\wss{Address Usability (\ppref{P_Usability}): Better documentation, with
attention on traceability, and automatic code generation are suggested when
designing software for change, and would help with these modifications.  So far
this problem of on-boarding new users has been addressed by updating the
documentation to better inform users of the underlying LBM theory and package
requirements. Similar issues with LBM parameters were noted by another
developer. Updating the user interface to better explain theoretical principles,
as well as test user input for compatibility, was the implemented solution. As
noted above, sometimes frequently asked questions on the underlying theory and
on how to use the software are answered in the documentation.  The interviewed
developer of ESPResSo commented that parts of their package's source code had
been refactored to Python to help address usability issues. Python was perceived
as a much more usable language, and it would be easy for future users and
developers to learn and understand the source code.}

\wss{Address Correctness (\ppref{P_Correctness}): Testing is not the only way to
build confidence in correctness.  Peer review.  }

\wss{Address quality of documentation (\ppref{P_Documentation}): Make required
background of users explicit, possibly in a requirements document.}
